Data Specialization for a Staged Lambda Calculus

### 2015-07-03 paper/figures/splitting-overview.tex
work on the splitting and semantics sections

\textbf{Splitting Terms at World \bbtwo\ ---
$\splittwo {\coltwo e A} A p l r$}
\begin{enumerate}
\item[] Precomputation $p$ is a monostaged term containing the first-stage computations in $e$.
\item[] Resumer $l.r$ is a monostaged term containing the second-stage computations in $e$.
\end{enumerate}

\textbf{Splitting Terms at World \bbonem\ ---
$\splitone {\colmix e A} A c {l.r}$}
\begin{enumerate}
\item[] Combined result $c$ is a monostage term containing the first-stage computations in $e$.
\item[] Resumer $l.r$ is a monostage term containing the second-stage computations in $e$.
\end{enumerate}

\textbf{Masking ---
$\rtab \xi v \vsplito \mval i q$}
\begin{enumerate}
\item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
\item[] Immediate value $i$ is a monostage value representing the first-stage components of $v$.
\item[] Residual $q$ is a monostaged term representing the second-stage computations in $\xi$ and $v$.
\end{enumerate}

\hrule
\vspace{1em}

\centering
\setlength{\unitlength}{2.8pt}
\begin{tabular}{c|c}
{\large \bf Splitting at World \bbonem} & 
{\large \bf Splitting at World \bbtwo} \\

\begin{picture} (80,54) (10,30)

\thicklines
\put(25,37){\oval(22,10)}
\put(25,75){\oval(22,10)}
\put(75,37){\oval(22,10)}
\put(75,75){\oval(22,10)}
\put(36,37){\vector(1,0){28}}
\put(36,75){\vector(1,0){28}}
\put(25,70){\vector(0,-1){28}}
\put(75,70){\vector(0,-1){28}}

\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\rtab \xi v$}}}
\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\mval i q$}}}

\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
\put(25,33){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\$\diaone e \xi v$}}

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produces the same result as 
	\lang\ evaluation followed by masking (left and bottom arrows),
	up to a term equivalence.
}}
\end{picture}

&
\begin{picture} (80,54) (10,30)

\thicklines
\put(25,75){\oval(22,10)}
\put(75,37){\oval(22,10)}
\put(75,75){\oval(22,10)}
\put(25,37){\vector(1,0){39}}
\put(36,75){\vector(1,0){28}}
\put(25,70){\line(0,-1){33}}
\put(75,70){\vector(0,-1){28}}

\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\coltwo e A$}}}
\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeS p l r$}}}
\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$q$}}}

\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\ $\diatwo e q$}}

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produces the same result as 
	\lang\ evaluation (left and bottom arrow),
	up to a term equivalence.
}}
\end{picture}

\\

\end{tabular}

\end{abstrsyn}
\caption{Overview of splitting.}
\label{fig:splittingSummary}
\end{figure*}

>>>>

\begin{theorem} \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

\begin{theorem}
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}

\begin{lemma}
\label{lem:twoCorrect}
If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonep e'$ and $\reduce {e'} i$, then $\reduce e i$.
\end{lemma}

% \textbf{Splitting Terms at World \bbtwo\ ---
% $\splittwo {\coltwo e A} A p l r$}
% \begin{enumerate}
% \item[] Precomputation $p$ is a monostaged term containing the first-stage computations in $e$.
% \item[] Resumer $l.r$ is a monostaged term containing the second-stage computations in $e$.
% \end{enumerate}

% \textbf{Splitting Terms at World \bbonem\ ---
% $\splitone {\colmix e A} A c {l.r}$}
% \begin{enumerate}
% \item[] Combined result $c$ is a monostage term containing the first-stage computations in $e$.
% \item[] Resumer $l.r$ is a monostage term containing the second-stage computations in $e$.
% \end{enumerate}

% \textbf{Masking ---
% $\rtab \xi v \vsplito \mval i q$}
% \begin{enumerate}
% \item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
% \item[] Immediate value $i$ is a monostage value representing the first-stage components of $v$.
% \item[] Residual $q$ is a monostaged term representing the second-stage computations in $\xi$ and $v$.
% \end{enumerate}

% \hrule
% \vspace{1em}

% \centering
% \setlength{\unitlength}{2.8pt}
% \begin{tabular}{c|c}
% {\large \bf Splitting at World \bbonem} & 
% {\large \bf Splitting at World \bbtwo} \\

% \begin{picture} (80,54) (10,30)

% \thicklines
% \put(25,37){\oval(22,10)}
% \put(25,75){\oval(22,10)}
% \put(75,37){\oval(22,10)}
% \put(75,75){\oval(22,10)}
% \put(36,37){\vector(1,0){28}}
% \put(36,75){\vector(1,0){28}}
% \put(25,70){\vector(0,-1){28}}
% \put(75,70){\vector(0,-1){28}}

% \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
% \put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\rtab \xi v$}}}
% \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
% \put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\mval i q$}}}

% \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
% \put(25,33){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
% \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
% \put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\$\diaone e \xi v$}}

% \put(28,66){\parbox[t]{44 \unitlength}{ 
% 	\textbf{Correctness Condition:} \\ 
% 	Splitting followed by evaluation (top and right arrows)
% 	produces the same result as 
% 	\lang\ evaluation followed by masking (left and bottom arrows),
% 	up to a term equivalence.
% }}
% \end{picture}

% &
% \begin{picture} (80,54) (10,30)

% \thicklines
% \put(25,75){\oval(22,10)}
% \put(75,37){\oval(22,10)}
% \put(75,75){\oval(22,10)}
% \put(25,37){\vector(1,0){39}}
% \put(36,75){\vector(1,0){28}}
% \put(25,70){\line(0,-1){33}}
% \put(75,70){\vector(0,-1){28}}

% \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\coltwo e A$}}}
% \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeS p l r$}}}
% \put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$q$}}}

% \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
% \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
% \put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\ $\diatwo e q$}}

% \put(28,66){\parbox[t]{44 \unitlength}{ 
% 	\textbf{Correctness Condition:} \\ 
% 	Splitting followed by evaluation (top and right arrows)
% 	produces the same result as 
% 	\lang\ evaluation (left and bottom arrow),
% 	up to a term equivalence.
% }}
% \end{picture}

% \\

% \end{tabular}

\end{abstrsyn}
\caption{\TODO put a real caption here and format the figure nicely}
\label{fig:splittingSummary}
\end{figure*}

### 2015-07-03 paper/semantics.tex
work on the splitting and semantics sections

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which includes and subsumes the concept of a stage.
Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms, 
with an underline (``$\exv v$'') noting the explicit inclusion of values into general terms.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
$\exv q$, and the second pass further reduces this residual to a value.
Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
Accordingly, this paper concerns itself only with the first pass of evaluation.

Since $\coltwo e A$ may be constructed out of terms at other world, 
our dynamics requires notions of values and steps that are specialized to each world.

@@

The statics and dynamics of \lang\ are related by a type safety theorem
comprising the usual progress and preservation lemmas, again annotated by world.
In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
\\{\bf Progress.~} 
\begin{itemize}
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
{\bf Preservation.~} 
\begin{itemize}
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}

@@

A multistage function, such as \texttt{qss} from \ref{sec:overview}, has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

>>>>

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which includes and subsumes the concept of a stage.
Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms, 
with underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
$\exv q$, and the second pass further reduces this residual.
Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
Consequently, this paper is concerned only with the first pass of evaluation.

Since $\coltwo e A$ may be constructed out of terms at other worlds, 
our dynamics requires notions of values and steps that are specialized to each world.

@@

The statics and dynamics of \lang\ are related by a type safety theorem
comprising progress and preservation, again annotated by world.
In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
\begin{theorem}\leavevmode
\label{thm:progress} 
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem}\leavevmode
\label{thm:preservation} 
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}

@@

The multistage function \texttt{qss} from \ref{sec:overview} has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}
